---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we load up the activity data and view a summary.

```{r}
library(plyr)
a <- data.frame(read.csv("activity.csv"))
summary(a)
```

Next we remove the NAs from the steps data.

```{r}
b <- a[!is.na(a$steps),]
summary(b)
```


## What is mean total number of steps taken per day?

The total daily steps are then calculated and a histogram is plotted.

```{r}
c <- aggregate(steps ~ date , data=b , sum)
hist(c$steps)
```

The mean and median number of steps per day are:

```{r}
mean(c$steps)
median(c$steps)
```

## What is the average daily activity pattern?

We then calculate another set giving the means accross the different intervals on all the days and plot a time series to show average steps across a day.

```{r}
d <- aggregate(steps ~ interval , data=b , mean)
plot(c(1:288), d$steps,type="l", xlab="Interval", ylab="Average Steps" )
```

The peak shown on the plot corresponds to the interval

```{r}
d[d$steps==max(d$steps),]$interval
```

Therefore the 5-minute interval 8:35 until 8:40.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
